---
title: "Simple workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TAPE)
```

The TAPE method in general takes a three-step framework:  

1. Adjust the phenotype for all individuals in the analysis based on the original case-control status and available family disease history;   
2. Fit null linear mixed model to get parameter estimates;  
3. Test for genetic association with the adjusted phenotype using score test statistics approximated by empirical saddlepoint approximation.  

This vignette illustrates the basic workflow for the null model estimation (Step 2) and testing procedure (Step 3) of TAPE. Note that the adjusted phenotypes generated from Step 1 are already included in the input dataframe for this example.  

#### 1. Input data  
Below is a minimal list of files needed for analysis using TAPE. Items 1-3 are for Step 2, Items 4-5 are for Step 3.  
``` r
# 1. Path to data frame including covariates and adjusted phenotype
file_cova = system.file("extdata", "example_cova.txt", package = "TAPE") 
# 2. Path to sparse matrix for close relatedness
file_kmat = system.file("extdata", "example_kmat.rds", package = "TAPE") 
# 3. Path to PLINK file to construct the Genetic relatedness matrix (also used for variance ratio calculation in example)
file_geno_grm = tools::file_path_sans_ext(system.file("extdata", "example_geno_grm.bim", package = "TAPE")) 

# 4. Genotype file for testing, if bgen format, include bgi file too
file_geno_test = tools::file_path_sans_ext(system.file("extdata", "example_geno_test.bgen", package = "TAPE"))
file_bgi_test = system.file("extdata", "example_geno_test.bgi", package = "TAPE")
# 5. Path to file with a column of IIDs in genotype file
file_idsingeno = system.file("extdata", "example_idsingeno.txt", package = "TAPE")
```

#### 2. Output Specification  
We then specify the destination for output files. Step 2 will generate a file of variance ratio information from sample markers and Step 3 will generate a summary file of test results.  
```{r}
file_output_s1 = paste0(dirname(file_cova),"/output_s1") # change this to your path for output file for step 2
file_output_s2 = paste0(dirname(file_cova),"/output_s2.txt") # change this to your path for output file for step 3
```

#### 3. `TAPE_Null_Model()`: Null model estimation (Step 2)   
We need to read in the covariate and phenotype data from `file_cova`, and (if using a 2-variance-component model) the sparse kinship matrix from `file_kmat`. The `K` argument in the function indicates the sparse kinship matrix, set `K=NULL` if sparse kinship information is not used in the mixed model.  
```{r}
data = read.table(file_cova,header=T)
kmat = readRDS(file_kmat)

# 2-variance-component
obj_null <- TAPE_Null_Model(y ~ sex+age, data = data, K=kmat, KgenFile=file_geno_grm, VRgenFile = file_geno_grm, idstoIncludeFile=file_idsingeno, tau=rep(0,3),fixtau=rep(0,3),outFile=file_output_s1, verbose=T)
# 2-variance-component LOCO
obj_null_loco <- TAPE_Null_Model(y ~ sex+age, data = data, K=kmat, KgenFile=file_geno_grm, VRgenFile = file_geno_grm, idstoIncludeFile=file_idsingeno, tau=rep(0,3),fixtau=rep(0,3),outFile=file_output_s1, verbose=T, loco=T)
# 1-variance-component (w/o sparse kinship )
obj_null_nok <- TAPE_Null_Model(y ~ sex+age, data = data, K=NULL, KgenFile=file_geno_grm, VRgenFile = file_geno_grm, idstoIncludeFile=file_idsingeno, tau=rep(0,2),fixtau=rep(0,2),outFile=file_output_s1, verbose=T)
```

#### 4. `TAPEtest()`: Score test (Step 3)  
In this step, the `null_object` argument takes the output object from previous step. Both bgen and PLINK binary format genotype files are supported as inputs. Set `genfile_format` argument to either "bgen" or "bfile" for the corresponding format. Note that if bgen format is used, the `bgi_file` option should also be specified.  
```{r}
n_variants_tested = TAPEtest(null_object=obj_null, genfile=file_geno_test, samplefile=file_idsingeno, outfile=file_output_s2, genfile_format="bgen", bgi_file="1")
```